Towards Generic Parallel Programming in Computer Science Education with Kokkos

### 2020-08-10 src/6.Backend.tex
Fix a few typos
Fix typos highlighted by the reviewers.

In this section we peak inside the Kokkos library and show how the a generic parallel for loop is mapped to the OpenMP back-end programming model. As a template metaprogramming library, Kokkos makes use of partial specialization. In particular the~\emph{parallel\_for} or ~\emph{parallel\_reduce} functional calls as shown in Figure~\ref{fig:KokkosExample} are mapped to partially specialized classes. Such classes correspond to particular back-end where for each back-end, concepts with partial specialization must be provided. Today, Kokkos supports the CUDA and OpenMP back-ends. For each new back-end, partially implemented classes that implement necessary concepts must be provided.

A simplified version of a partially specialized class for a parallel for loop implementation is shown in Figure\ref{fig:KokkosExampleOMPBackEnd}. The ~\emph{parallel\_for} function instantiates a partial specialization of the ~\emph{ParallelFor} class and calls its ~\emph{execute} function. That function is then responsible to implement the Kokkos parallel pattern. In this back-end, the parallel implementation is generated by using OpenMP pragma annotations and a by invoking a compatible compiler when the library is compiled. 
\begin{figure}
\begin{small}
\begin{minted}[fontsize=\small]{c++}
template <class FunctorType>
class ParallelFor<FunctorType,
              RangePolicy<OpenMP>> 
{
  const FunctorType functor;
  const RangePolicy<OpenMP> policy; 
public:
  void execute() const {
    #pragma omp parallel for
    for(int i=policy.begin();
            i<policy.end();
            ++i)
      functor(i);
  }
};
template <class FunctorType,class PolicyType>
void parallel_for(std::string label, 
    const PolicyType& policy,
    const FunctorType& functor) {
  // Call profiling tool hooks with label
  ParallelFor<FunctorType,PolicyType> 
    pf{functor,policy};
  pf.execute();
}
\end{minted}
\end{small}
\caption{This code shows the back-end implementation of a parallel loop in the Kokkos programing model. It shows a particular implementation of a generic pattern as a templated class to the OpenMP execution space. Also it shows the use of OpenMP pragma annotations that put platform compilers in charge to generate parallel code for the underlying hardware architecture.}
\label{fig:KokkosExampleOMPBackEnd}
\end{figure}

%\begin{figure}
%\centerline{\includegraphics[width=0.3\textwidth]{img/Build.png}}
%\caption{The compilation workflow includes the invocation of the platform compiler that is in charge to generate parallel   Building workflow}
%\label{fig:workflow}  
%\end{figure}

>>>>

In this section we peek inside the Kokkos library and show how a generic parallel for loop is mapped to the OpenMP backend programming model. As a template metaprogramming library, Kokkos makes use of partial specialization. In particular the~\emph{parallel\_for} or ~\emph{parallel\_reduce} functional calls as shown in Figure~\ref{fig:KokkosExample} are mapped to partially specialized classes. Such classes correspond to a particular back-end where for each back-end, concepts with partial specialization must be provided. Today, Kokkos supports the CUDA, OpenMP, HIP as well as one-sided MPI, SHMEM and NVSHMEM back-ends. For each new back-end, partially implemented classes that implement necessary concepts must be provided.

A simplified version of a partially specialized class for a parallel for loop implementation is shown in Figure\ref{fig:KokkosExampleOMPBackEnd}. The ~\emph{parallel\_for} function instantiates a partial specialization of the ~\emph{ParallelFor} class and calls its ~\emph{execute} function. That function is then responsible to implement the Kokkos parallel pattern. In this back-end, the parallel implementation is generated by using OpenMP pragma annotations and a by invoking a compatible compiler when the library is compiled. 
\begin{figure}
\begin{small}
\begin{minted}[fontsize=\small]{c++}
template <class FunctorType>
class ParallelFor<FunctorType,
              RangePolicy<OpenMP>> 
{
  const FunctorType functor;
  const RangePolicy<OpenMP> policy; 
public:
  void execute() const {
    #pragma omp parallel for
    for(int i=policy.begin();
            i<policy.end();
            ++i)
      functor(i);
  }
};
template <class FunctorType,class PolicyType>
void parallel_for(std::string label, 
    const PolicyType& policy,
    const FunctorType& functor) {
  // Call profiling tool hooks with label
  ParallelFor<FunctorType,PolicyType> 
    pf{functor,policy};
  pf.execute();
}
\end{minted}
\end{small}
\caption{This code shows the back-end implementation of a parallel loop in the Kokkos programming model. It shows a particular implementation of a generic pattern as a templated class to the OpenMP execution space. Also, it shows the use of OpenMP pragma annotations that put platform compilers in charge to generate parallel code for the underlying hardware architecture.}
\label{fig:KokkosExampleOMPBackEnd}
\end{figure}

### 2020-08-10 src/8.Related.tex
Fix typos highlighted by the reviewers.

C++ increasingly supports parallel constructs. It is becoming increasingly possible to teach parallelism in C++ and relying on modern C++ compilers. However, at time of publication, C++ does not support accelerators, or many of the complexities of parallel computing related to accelerator programming. In courses that do not target accelerators, this can be a relevant option.

@@

RAJA~\cite{RAJA} is a C++ abstraction library with similarities to Kokkos. The important different however is that Kokkos is descriptive, while RAJA is prescriptive. In Kokkos, the programming models determines how an application is mapped to the underlying hardware. RAJA provides functionality that exposes hardware details and relies on the developer to try different strategies to map kernels to architectures. RAJA is a viable option for class education in which the educator would like to expose students to the semantics of different programming models without exposing them to the detailed syntax of those models.

DPC++\cite{DPCPP} is a model developed by Intel\textsuperscript{\textregistered} for expressing parallelism across Intel architectures. In addition to sharing a descriptive philosophy with Kokkos, it also shares parallel patterns. Unlike Kokkos, at time of publication it targets Intel architectures only and is not vendor-agnostic. A course targeting Intel architectures, DPC++ could be a valid option.

>>>>

C++ increasingly supports parallel constructs. It is becoming increasingly possible to teach parallelism in C++ and relying on modern C++ compilers. However, at the time of publication, C++ does not support accelerators or many of the complexities of parallel computing related to accelerator programming. In courses that do not target accelerators, this can be a relevant option.

@@

RAJA~\cite{RAJA} is a C++ abstraction library with similarities to Kokkos. The important difference however is that Kokkos is descriptive, while RAJA is prescriptive. In Kokkos, the programming models determine how an application is mapped to the underlying hardware. RAJA provides functionality that exposes hardware details and relies on the developer to try different strategies to map kernels to architectures. RAJA is a viable option for class education in which the educator would like to expose students to the semantics of different programming models without exposing them to the detailed syntax of those models.

DPC++\cite{DPCPP} is a model developed by Intel\textsuperscript{\textregistered} for expressing parallelism across Intel architectures. In addition to sharing a descriptive philosophy with Kokkos, it also shares parallel patterns. Unlike Kokkos, it targets Intel architectures only and is not vendor-agnostic. A course targeting Intel architectures, DPC++ could be a valid option.

### 2020-08-10 src/9.Conclusion.tex
Fix typos highlighted by the reviewers.

\section{Conclusion}\label{chap:conclusion}
In this work we gave an introduction to the Kokkos parallel programming model and presented guiding thoughts behind its design. We discussed a machine representation of a parallel hardware architecture that is characterized by a variety of processing units, memories and their hierarchical organization. This machine model allowed us to define a constrained semantic represented though different types and patterns. Further we listed supported paradigms and considerations that influenced the design of the semantic capture.

@@

To the educational community, we presented Kokkos as a representative programming model of a trend towards support of parallel programming in base languages and libraries. We believe in the importance of embracing this trend in computer science education as such parallel programming models can contribute to the generic and relevant education of parallel programming and prepare student for their programming careers.

Finally we would like to express gratitude to educators for taking on this important role. We are open to provide further material or support and propose the interested reader to reach out to the development team on Slack\cite{KOKKOS_SLACK}.

>>>>

\section{Conclusion}\label{chap:conclusion}
In this work we gave an introduction to the Kokkos parallel programming model and presented guiding thoughts behind its design. We discussed a machine representation of a parallel hardware architecture that is characterized by a variety of processing units, memories, and their hierarchical organization. This machine model allowed us to define a constrained semantic represented though different types and patterns. Further, we listed supported paradigms and considerations that influenced the design of the semantic capture.

@@

To the educational community, we presented Kokkos as a representative programming model of a trend towards the support of parallel programming in base languages and libraries. We believe in the importance of embracing this trend in computer science education as such parallel programming models can contribute to the generic and relevant education of parallel programming and prepare students for their programming careers.

Finally, we would like to express gratitude to educators for taking on this important role. We are open to provide further material or support and propose the interested reader to reach out to the development team on Slack\cite{KOKKOS_SLACK}.
